---
title: "p8105_mtp_jy3269"
author: "Jingyi Yao"
date: "`r Sys.Date()`"
output: github_document
---

```{r,warning=FALSE, message=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)
library(readxl)
wordcountaddin::text_stats("p8105_mtp_jy3269.Rmd")

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```

##  Data

**Introduction to the Raw Data:**

  * The data records 3 pet dogs' weight from `r weight$date[1]` to `r weight$date[-1]`. 

  * Raw dataset has `r nrows(raw_data)` rows and 6 columns.

  * The first column is the date when the dogs are weighed. 

  * The next 3 columns show the weight of each dog, named Raisin, Simone and Gagne respectively.

  * A standard object is also weighed and recorded occasionally as a reference. 

  * The last column contains occasional notes for the dogs' status.


**Goals of the Report:**

  * Show the major steps in turning the raw data set into a tidy one available for analysis.
  * Explore the features in the 3 dogs' weight and try to interpret them.
  * Visualize the distribution of dogs' weights and identify their variation over time.


### I. Weight 

#### Issue 1: Incorrect column names  

The raw data is imported from excel. Because of the merged cells, the column names are not correct in `raw_data`.

```{r}
raw_data <- read_excel("./data/dogweights_07may2022.xlsx",range = "B1:F363") 
head(raw_data)    # show part of the raw data
```

Extract the first row as column names to rename it. Then, remove the first row to avoid duplication.
```{r}
weight = raw_data             # save the raw data and work on the new data set named weight 

colnames(weight) = weight[1,] # change column names

weight <- weight[-1,]         # omit the first row which is the column name

head(weight)                  # data after column name correction
```


#### Issue 2: Unify Units

The original units for weight contain both lbs and oz. First, the two units are separated. According to 1 lb = 16 oz, oz is transformed into lb and then added to the lbs column. Now we have the weight measured in lbs.
```{r}
weight <- weight %>% 
  separate(Raisin, c("Raisin (Lbs)","Raisin (Oz)"), sep = " ", convert = TRUE) %>% 
  separate(Simone, c("Simone (Lbs)","Simone (Oz)"), sep = " ", convert = TRUE) %>% 
  separate(Gagne, c("Gagne (Lbs)","Gagne (Oz)"), sep = " ", convert = TRUE) %>% 
  separate(std, c("std (Lbs)","std (Oz)"), sep = " ", convert = TRUE) %>% 
  janitor::clean_names() %>% 
  mutate(Raisin = raisin_lbs + raisin_oz/16,   # 1 lb = 16 oz
         Simone = simone_lbs + simone_oz/16,
         Gagne = gagne_lbs + gagne_oz/16,
         Standard = std_lbs + std_oz/16
          ) %>% 
  select(date,Raisin,Simone,Gagne,Standard)
```

```{r}
head(weight)     # data after units transformation
```


#### Issue 3: Messy date shown in R

1. The dates imported from excel are shown as integers but in character type in R. 
   First, transform it into numeric type. Then, add the number to the original date of `1899-12-30` to get the actual date.
   
2. A fault in date.
   The fault is detected when examining the unique values of dates. There is an `s` in the date column. We suppose that this is a wrong entry due to mistyping. We locate it and use the middle day of the 2 dates next to it to correct it.

```{r}
unique(weight$date)                # find outliers in date

which(weight$date == "s")          # locate the outlier
```

```{r}
weight <- weight %>% 
  mutate(date = as.numeric(date))  # transform data type and the “s” will be NA

# use the middle day between the 76th and 78th day.
weight$date[is.na(weight$date)] <- round(((weight$date)[78] + (weight$date)[76])/2)
```


```{r}
# change the numbers into date format
weight <- weight %>% 
  mutate(date = as.Date(weight$date, origin = "1899-12-30"))

write_csv(weight,file = "data/weight.csv")

# another data set to store the date in a separated way
weight_1 <- weight %>%  
  separate(date, into = c("Year","Month","Day"),sep = "-") %>% 
  mutate(Month = month.abb[as.numeric(Month)])
```


```{r}
head(weight)    # data after date correction
```



#### 4. Tidy data 

Transform the 4 columns of weight into 2 new columns named `Dog` and `Weight`. 

  * `Dog` contains dogs' and object's names.
  * `Weight` stores the weight with respect to the name.
```{r}
weight_2 <- weight %>% 
  pivot_longer(                          
    Raisin:Standard,
    names_to = "Dog", 
    values_to = "Weight") %>% 
  mutate(
    Dog = factor(Dog),
  )

weight_2
```



### II. Notes

Extract the last column from raw data and bind with the tidied date column to form the `note` dataset
```{r}
raw_data_note <- read_excel("./data/dogweights_07may2022.xlsx",range = "G2:G363") 
raw_data_note

date <- weight[,1] # use the date in the weight dataframe
date

note <- bind_cols(date,raw_data_note)

colnames(note)[2] <- "Notes"

note

write_csv(note,file = "data/note.csv")
```



##  EDA

#### 1.unique date

there are `r length(unique(weight$date))` unique dates.
```{r echo=FALSE}
num_unique_date <- length(unique(weight$date))

```


#### 2. Make a table showing number of observations, mean and standard deviation of weight
```{r}
weight_2 %>% 
  group_by(Dog) %>% 
  summarize(
    avg_weight = round(mean(Weight,na.rm = T),digits = 3),
    sd_weight = round(sd(Weight,na.rm = T),digits = 3),
    num_obs = n()
  ) %>% 
  knitr::kable(caption = "Weight measured in lbs")


```



## Visualization

### right plot -- Each Dog’s Weight (lbs) Over Time
```{r}
weight_lbs_over_time <- weight_2 %>% 
  filter(Dog != "Standard") %>% 
  ggplot(aes(x = date, y = Weight,color = Dog), group = Dog) +
  geom_line(size = 0.8) + facet_wrap(.~Dog, nrow=3,scales = "free_y")+
  labs(x = "Time",
       y = "Weight  (lbs)",
       title = "Each Dog’s Weight (lbs) Over Time") +
  theme_bw() +
  theme(legend.position = "bottom")

weight_lbs_over_time
  
```




### left plot -- distribution of weights (lbs) for each dog

```{r}
weight_distribution_lbs<- weight_2 %>% 
  filter(Dog != "Standard") %>%
  ggplot(aes(x = Dog, y = Weight,fill = Dog)) +
  geom_violin() +
  scale_y_continuous(
    limits = c(6,20)) +
  labs(x = "Dog",
       y = "Weight  (lbs)",
       title = "Weight Distribution (lbs) of Each Dog") +
  theme_bw() +
  theme(legend.position = "bottom")
  

weight_distribution_lbs
```


### combine the 2 plots

```{r}

weight_distribution_lbs + weight_lbs_over_time

```




