---
title: "p8105_mtp_jy3269"
author: "Jingyi Yao"
date: "`r Sys.Date()`"
output: github_document
---

```{r,warning=FALSE, message=FALSE,echo=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)
library(readxl)
wordcountaddin::text_stats("p8105_mtp_jy3269.Rmd")

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```

##  Data

**Introduction:**

  * The data records 3 pet dogs' weight from 12-Aug-18 to 6-May-22. 

  * It has 362 rows and 6 columns.

  * The first column is the date. 

  * The next 3 columns show the weight of each dog, named Raisin, Simone and Gagne respectively.

  * A standard object is also weighed occasionally as a reference. 

  * The last column contains occasional notes for the dogs' status.


**Goals of the Report:**

  * Show the major steps in turning the raw data set into a tidy one
  * Explore the features in the 3 dogs' weight and try to interpret them
  * Visualize the distribution of dogs' weights and identify their variation over time


### I. Weight 

#### 1: Incorrect column names  

The column names are not correct in `raw_data`.

```{r,echo=FALSE}
raw_data <- read_excel("./data/dogweights_07may2022.xlsx",range = "B1:F363") 
head(raw_data)    # show part of the raw data
```

Extract the first row as column names to rename it. Then, remove the first row to avoid duplication.
```{r}
weight = raw_data             # save the raw data and work on the new data set named weight 

colnames(weight) = weight[1,] # change column names

weight <- weight[-1,]         # omit the first row which is the column name

head(weight)                  # data after column name correction
```


#### 2: Unify Units

The original units contain both lbs and oz. First, the two units are separated. According to 1 lb = 16 oz, oz is transformed into lb and then added to the lbs column. Now we have the weight measured in lbs.
```{r}
weight <- weight %>% 
  separate(Raisin, c("Raisin (Lbs)","Raisin (Oz)"), sep = " ", convert = TRUE) %>% 
  separate(Simone, c("Simone (Lbs)","Simone (Oz)"), sep = " ", convert = TRUE) %>% 
  separate(Gagne, c("Gagne (Lbs)","Gagne (Oz)"), sep = " ", convert = TRUE) %>% 
  separate(std, c("std (Lbs)","std (Oz)"), sep = " ", convert = TRUE) %>% 
  janitor::clean_names() %>% 
  mutate(Raisin = raisin_lbs + raisin_oz/16,   # 1 lb = 16 oz
         Simone = simone_lbs + simone_oz/16,
         Gagne = gagne_lbs + gagne_oz/16,
         Standard = std_lbs + std_oz/16
          ) %>% 
  select(date,Raisin,Simone,Gagne,Standard)
```

```{r}
head(weight)     # data after units transformation

```


#### 3: Messy date 

1. The dates from excel are shown as integers but in character type in R. 
   First, transform it into numeric type. Then add the number to the original date of `1899-12-30` to get the actual date
   
2. Wrong date 
   When examining the unique values of dates,there is an `s` in the date column. We suppose that this is a wrong entry due to mistyping. We locate it and use the middle day of the 2 dates next to it to correct it

```{r,eval = FALSE}
unique(weight$date)                # find outliers in date

which(weight$date == "s")          # locate the outlier
```

```{r,warning=FALSE}
weight <- weight %>% 
  mutate(date = as.numeric(date))  # transform data type and the “s” will be NA

# use the middle day between the 76th and 78th day.
weight$date[is.na(weight$date)] <- round(((weight$date)[78] + (weight$date)[76])/2)
```


```{r}
# change the numbers into date format
weight <- weight %>% 
  mutate(date = as.Date(weight$date, origin = "1899-12-30"))

write_csv(weight,file = "data/weight.csv")

# another data set to store the date in a separated way
weight_1 <- weight %>%  
  separate(date, into = c("Year","Month","Day"),sep = "-") %>% 
  mutate(Month = month.abb[as.numeric(Month)])
```


```{r}
head(weight)    # data after date correction
```


#### 4. Wrong Weight

When examine the unique values of weights, we find wrong entry like `1810` instead of `18 10`.

```{r}
weight$Raisin[40] <- 18 + 10/16
weight$Raisin[117] <- 18 + 3/16
weight$Raisin[210] <- 17 + 15/16
weight$Standard[358] <- 20 + 15.4/16
```



#### 5. Tidy data 

Transform columns of weight into 2 new columns named `Dog` and `Weight`. 
```{r}
weight_2 <- weight %>% 
  pivot_longer(                          
    Raisin:Standard,
    names_to = "Dog", 
    values_to = "Weight") %>% 
  mutate(
    Dog = factor(Dog),
  )

weight_2
```



### II. Notes

Extract last column and bind with the date 
```{r}
raw_data_note <- read_excel("./data/dogweights_07may2022.xlsx",range = "G2:G363") 
raw_data_note

date <- weight[,1] # use the date in the weight dataframe
date

note <- bind_cols(date,raw_data_note)

colnames(note)[2] <- "Notes"

note

write_csv(note,file = "data/note.csv")
```



##  EDA

#### 1.unique dates

`r length(unique(weight$date))` unique dates.
```{r echo=FALSE}
num_unique_date <- length(unique(weight$date))

```


#### 2. Table

  * `avg_weight` measures the average weight. Simone is the lightest and Raisin is the heaviest.
  * `sd_weight` measures the deviation from the average weight. The deviations are all small ,showing low variability in weight. Variability of Raisin and Simone are slightly higher.
  * `num_obs` counts the non-NA observations.
```{r}
weight_2 %>% 
  group_by(Dog) %>% 
  summarize(
    avg_weight = round(mean(Weight,na.rm = T),digits = 3),   # NA is removed
    sd_weight = round(sd(Weight,na.rm = T),digits = 3),
    num_obs = sum(!is.na(Weight))                            # count non-NA obs
  ) %>% 
  knitr::kable(caption = "Number of Observations and Weight Features")


```



## Visualization

```{r,warning=FALSE}
weight_lbs_over_time <- weight_2 %>% 
  filter(Dog != "Standard") %>% 
  ggplot(aes(x = date, y = Weight,color = Dog), group = Dog) +
  geom_line(size = 0.8) + facet_wrap(.~Dog, nrow=3,scales = "free_y")+
  labs(x = "Time",
       y = "Weight  (lbs)",
       title = "Each Dog’s Weight (lbs) Over Time") +
  theme_bw() +
  theme(legend.position = "bottom")

weight_lbs_over_time
  
```


  * Gagne's observations ended in 2019 because it died on 09/08/2019
  * A sudden drop in Raisin's weight is on 08/14/2019. The value recorded is 7lbs 15oz, while the weight on 08/10/2019 is 17lbs 15oz. Thus, this outlier is probably caused by mistyping.
  * The interrupted connections are due to no records. Most of the missing records are due to that dogs are at the kennels.
  * The overall trends of Gagne and Simone are decreasing while Raisin has a more stabilized trend excluding the outlier.



```{r}
weight_distribution_lbs<- weight_2 %>% 
  filter(Dog != "Standard") %>%
  ggplot(aes(x = Dog, y = Weight,fill = Dog)) +
  geom_violin() +
  scale_y_continuous(
    limits = c(6,20)) +
  labs(x = "Dog",
       y = "Weight  (lbs)",
       title = "Weight Distribution (lbs) of Each Dog") +
  theme_bw() +
  theme(legend.position = "bottom")
  

weight_distribution_lbs
```

  * Gagne and Simone have similar distribution, ranging from 5 to 12 lbs. And most of the observations concentrate around 11 lbs.
  * Raisin's weights mostly concentrate around 18 lbs. And the lower range of 7 lbs could be the outlier caused by mistyping.


```{r}

weight_distribution_lbs + weight_lbs_over_time

```




